---
title: "ch3-lab-linear-regression"
author: "Kurt Schuepfer"
date: "1/29/2017"
output: html_document
---
#Chapter 3 Lab: Linear Regression Lab

####3.6.1

```{r}
library(MASS)
library(ISLR)
```

####3.6.2 Simple Linear Regression
```{r, echo = TRUE}
attach(Boston)
names(Boston)
lm.fit <- lm(medv~lstat)
summary(lm.fit)
coefficients(lm.fit)
confint(lm.fit)

#can then use the regresion to predict unknown data apply confidence or prediction intervals 
predict(lm.fit, data.frame(lstat = c(5, 10, 15)), interval = "confidence")
predict(lm.fit, data.frame(lstat = c(5, 10, 15)), interval = "prediction")

plot(lstat, medv)
abline(lm.fit, lwd = 3, col = "red", pch = 1:20)
par(mfrow = c(2,2))
plot(lm.fit)
plot(predict(lm.fit), residuals(lm.fit))
plot(predict(lm.fit), rstudent(lm.fit)) ##standardized residuals

plot(hatvalues(lm.fit)) #leverage statistics for any value
which.max(hatvalues(lm.fit))
```

####3.6.3 Multiple Linear Regression
```{r}
lm.fit <- lm(medv~lstat+age, data=Boston)
summary(lm.fit)
lm.fit <- lm(medv~., data = Boston)
summary(lm.fit)
summary(lm.fit)$r.sq
summary(lm.fit)$sigma

#VIF - variance inflation factor
library(car)
vif(lm.fit)

##removing one predictor
lm.fit1 <- lm(medv~.-age, data = Boston)
summary(lm.fit1)
lm.fit1 = update(lm.fit, ~.-age) #update function is an alternative
```

####3.6.4 Interaction Terms
```{r}
summary(lm(medv~lstat*age, data = Boston))
```
####3.6.5 Non-linear Transformations of the Predictors

Must use I() function to fit quadratic terms or higher-order polynomials
```{r}
lm.fit <- lm(medv~lstat) #linear submodel
lm.fit2 <- lm(medv~lstat + I(lstat^2))
summary(lm.fit2)

##Model with quadratic term is highly significant. Now compare to previous fit.
anova(lm.fit, lm.fit2) #null is that models fit equally well
par(mfrow=c(2,2))
plot(lm.fit2)
```

Can use poly() function to fit highe-rorder polynomials
```{r}
lm.fit5 <- lm(medv~poly(lstat, 5))
summary(lm.fit5)
```

Can also transform in other ways
```{r}
summary(lm(medv~log(rm), data = Boston))
```

####3.6.6 Qualitative Predictors
```{r}
attach(Carseats)
names(Carseats)
```

R automatically generates dummy variables for factor variables

```{r}
lm.fit <- lm(Sales~.+Income:Advertising+Price:Age, data = Carseats)
summary(lm.fit)
```

```{r}
contrasts(ShelveLoc) #shows dummy coding schemes for variables
?contrasts 
```

####3.6.7 Writing Functions

```{r}
LoadLibraries = function(){
        library(ISLR)
        library(MASS)
        print("The libraries have been loaded.")
}

LoadLibraries()
LoadLibraries
```

