---
title: "ch4-lab-classification"
author: "Kurt Schuepfer"
date: "2/7/2017"
output: html_document
---
#Chapter 4 Lab: Classification

LDA is like PCA but it focuses on maximizing the seperability between known categories 
https://www.youtube.com/watch?v=azXCzI57Yfc


Puts multiple dimensions onto a single dimension, maximizing distance among categories while minimizing scatter (variability within each category)
 - We want big mean differences and minimal scatter
 - If 3 categories, find a point that is central to the data and then measure the distances between each category's centroid and the central point
 
Similar to PCA, except PCA does not try to maximize distance among categories. It just looks to slice the data where there is the most variation. 

Similarities to PCA
 - Both methods rank the new axes they create in order of importance
 - PC1 accounts for hte most variation in the data
 - PC2 does the second best job, etc.
 - LD1 accounts for the most variation between the categories
 - LD2 does the second best job, etc. 


####4.

```{r}
library(MASS)
library(ISLR)
```

####4.6.1 Stock Market Data

```{r}
names(Smarket)
summary(Smarket)
cor(Smarket)
cor(Smarket[,-9])
```

As one would expect, the correlations between the lag variables and to- day’s returns are close to zero. In other words, there appears to be little correlation between today’s returns and previous days’ returns. The only substantial correlation is between Year and Volume. By plotting the data we see that Volume is increasing over time. In other words, the average number of

```{r}
attach(Smarket)
plot(Volume)
```

####4.6.2 Logistic Regression

Next, we will fit a logistic regression model in order to predict Direction using Lag1 through Lag5 and Volume.The glm() function fits generalized glm() linear models, a class ofmodels that includes logistic regression. The syntax of the glm() function is similar to that of lm(), except that we must pass in the argument family=binomial in order to tell R to run a logistic regression rather than some other type of generalized linear model

```{r}
glm.fit <- glm(Direction~Lag1+Lag2+Lag3+Lag4+Lag5+Volume, data = Smarket, family = binomial)
summary(glm.fit)
coef(glm.fit)
```

The predict() function can be used to predict the probability that the market will go up, given values of the predictors. The type="response"
option tells R to output probabilities of the form P(Y =1|X), as opposed to other information such as the logit. If no data set is supplied to the predict() function, then the probabilities are computed for the training data that was used to fit the logistic regression model. Here we have printed only the first ten probabilities. We know that these values correspond to the probability of the market going up, rather than down, because the contrasts() function indicates that R has created a dummy variable with a1 for Up.

```{r}
glm.probs <- predict(glm.fit, type = "response")
glm.probs[1:10]
contrasts(Direction)
```

In order to make a prediction as to whether the market will go up or down on a particular day, we must convert these predicted probabilities into class labels, Up or Down. The following two commands create a vector of class predictions based on whether the predicted probability of a market increase is greaterthan orless than 0.5.

```{r}
glm.pred <- rep("Down", 1250)
glm.pred[glm.probs > .5] = "Up"
```

The first command creates a vector of 1,250 Down elements. The second line transforms to Up all of the elements for which the predicted probability of a
market increase exceeds 0.5. Given these predictions, the table() function table() can be used to produce a confusion matrix in order to determine how many observations were correctly or incorrectly classified

```{r}
table(glm.pred, Smarket$Direction)
mean(glm.pred == Smarket$Direction)
```

In this case, logistic regression correctly predicted the movement of the market 52.2% of the time.

Try on a validation set
```{r}
train <- Smarket$Year<2005
Smarket.2005 <- Smarket[!train,]
dim(Smarket.2005)
Direction.2005 <- Smarket$Direction[!train]
```

We now fit a logistic regression model using only the subset of the obser-
vations that correspond to dates before 2005, using the subset argument. We then obtain predicted probabilities of the stock market going up for each of the days in our test set—that is, for the days in 2005.

```{r}
glm.fit <- glm(Direction~Lag1+Lag2+Lag3+Lag4+Lag5+Volume,
               data = Smarket, family=binomial, subset=train)
glm.probs <- predict(glm.fit, Smarket.2005, type = "response")
```

Notice that we have trained and tested our model on two completely sep- arate data sets: training was performed using only the dates before 2005, and testing was performed using only the dates in 2005. Finally, we com- pute the predictions for 2005 and compare them to the actual movements of the market over that time period.

```{r}
glm.pred <- rep("Down", 252)
glm.pred[glm.probs>.5]="Up"
table(glm.pred, Direction.2005)
mean(glm.pred==Direction.2005)
mean(glm.pred!=Direction.2005)
```
The results are rather disappointing: the test error rate is 52%, which is worse than random guessing! Of course this result is not all that surprising, given that one would not generally expect to be able to use previous days’ returns to predict future market performance. (After all, if it were possible to do so, then the authors of this book would be out striking it rich rather than writing a statistics textbook.)

We recall that the logistic regression model had very underwhelming p-
values associated with all of the predictors, and that the smallest p-value, though not very small, corresponded to Lag1. Perhaps by removing the variables that appear not to be helpful in predicting Direction,we can obtain a more effective model. After all, using predictors that have no relationship with the response tends to cause a deterioration in the test error rate (since such predictors cause an increase in variance without a corresponding decrease in bias), and so removing such predictors may in turn yield an improvement

```{r}
glm.fit <- glm(Direction~Lag1+Lag2, data = Smarket, 
               family=binomial, subset=train)
glm.probs <- predict(glm.fit, Smarket.2005, type="response")
glm.pred=rep("Down", 252)
glm.pred[glm.probs>.5]="Up"
table(glm.pred, Direction.2005)
mean(glm.pred==Direction.2005)
```
Now the results appear to be a little better: 56% of the daily movements have been correctly predicted. It is worth noting that in this case, a much simpler strategy of predicting that the market will increase every day will also be correct 56% of the time! Hence, in terms of overall error rate, the logistic regression method is no better than the na¨
ıve approach. However,
the confusion matrix shows that on days when logistic regression predicts an increase in the market, it has a 58% accuracy rate. This suggests a possible trading strategy of buying on days when the model predicts an in- creasing market, and avoiding trades on days when a decrease is predicted. Of course one would need to investigate more carefully whether this small improvement was real or just due to random chance.

Suppose that we want to predict the returns associated with particular
values of Lag1 and Lag2. In particular, we want to predict Direction on a day when Lag1 and Lag2 equal 1.2 and 1.1, respectively, and on a day when they equal 1.5 and −0.8. We do this using the predict() function.

```{r}
predict(glm.fit, newdata=data.frame(Lag1=c(1.2, 1.5), Lag2=c(1.1,-.8)),
        type="response")
```

####4.6.3 Linear Discriminant Analysis

Now we will perform LDA on the Smarket data. In R,we fitanLDA model using the lda() function, which is part of the MASS library. Notice that the lda() syntax for the lda() function is identical to that of lm(), and to that of
glm() except for the absence of the family option. We fit the model using only the observations before 2005.

```{r}
library(MASS)
lda.fit <- lda(Direction~Lag1+Lag2, data = Smarket, subset=train)
lda.fit
```

Prior probs show that 49.2% of training observations correspond to days during which the market went down. 

Group means are average of each predictor within each class
Here, suggests that there is a tendency for the previous 2 days’ returns to be negative on days when the market increases, and a tendency for the previous days’ returns to be positive on days when the market declines.

The coefficients of linear discriminants output provides the linear combination of Lag1 and Lag2 that are used to form the LDA decision rule. In other words, these are the multipliers of the elements of X = x in (4.19). If −0.642×Lag1−0.514×Lag2 is large, then the LDA classifier will predict a market increase, and if it is small, then the LDA classifier will predict a market decline. 

The plot() function produces plots of the linear
discriminants, obtained by computing −0.642 × Lag1 − 0.514 × Lag2 for each of the training observations.

```{r}
lda.pred <- predict(lda.fit, Smarket.2005)
names(lda.pred)
```
As we observed in Section 4.5, the LDA and logistic regression predictions are almost identical.

```{r}
lda.class <- lda.pred$class
table(lda.class, Direction.2005)
mean(lda.class==Direction.2005)
```

Applying a 50% threshold to the posterior probabilities allows us to recre- ate the predictions contained in lda.pred$class.

```{r}
sum(lda.pred$posterior[,1] >= .5)
sum(lda.pred$posterior[,1] < .5)
```
Notice that the posterior probability output by the model corresponds to the probability that the market will decrease:

```{r}
lda.pred$posterior[1:20,1]
lda.class[1:20]
```
If we wanted to use a posterior probability threshold other than 50% in order to make predictions, then we could easily do so. For instance, suppose that we wish to predict a market decrease only if we are very certain that the market will indeed decrease on that day—say, if the posterior probability is at least 90%.

```{r}
sum(lda.pred$posterior[,1] >= .9)
```
No days in 2005 meet that threshold! In fact, the greatest posterior prob- ability of decrease in all of 2005 was 52.02%.

####4.6.3 Quadratic Discriminant Analysis

We will now fit a QDA model to the Smarket data. QDA is implemented in R using the qda() function, which is also part of the MASS library. The qda() syntax is identical to that of lda().

```{r}
qda.fit <- qda(Direction~Lag1+Lag2, data = Smarket, subset=train)
qda.fit
```

Note same output of LDA, except no section for coefficients of linear discriminants:

Because the QDA classifier involves a quadratic, rather than a linear, function of the predictors.

```{r}
qda.class=predict(qda.fit, Smarket.2005)$class
table(qda.class, Direction.2005)
mean(qda.class==Direction.2005)
```
Interestingly, the QDA predictions are accurate almost 60% of the time, even though the 2005 data was not used to fit the model. This level of accu- racy is quite impressive for stock market data, which is known to be quite hard to model accurately. This suggests that the quadratic form assumed by QDA may capture the true relationship more accurately than the linear forms assumed by LDA and logistic regression. However, we recommend evaluating this method’s performance on a larger test set before betting that this approach will consistently beat the market!

####4.6.5 K-Nearest Neighbors

We will now perform KNN using the knn() function, which is part of the knn() class library. This function works rather differently from the other model- fitting functions that we have encountered thus far. Rather than a two-step approach in which we first fit the model and then we use the model to make predictions, knn() forms predictions using a single command. The function requires four inputs.

1. A matrix containing the predictors associated with the training data, labeled train.X below.
2. Amatrix containing the predictors associated with the data for which we wish to make predictions, labeled test.X below.
3. A vector containing the class labels for the training observations, labeled train.Direction below.
4. A value for K, the number of nearest neighbors to be used by the classifier

```{r}
library(class)
train.X = cbind(Lag1,Lag2)[train,]
test.X <- cbind(Lag1,Lag2)[!train,]
train.Direction = Direction[train]
```

```{r}
set.seed(1)
knn.pred = knn(train.X, test.X, train.Direction, k=1)
table(knn.pred, Direction.2005)
```
The results using K = 1 are not very good, since only 50% of the observa- tions are correctly predicted. Of course, it may be that K = 1 results in an overly flexible fit to the data. Below, we repeat the analysis using K =3.

```{r}
knn.pred <- knn(train.X, test.X, train.Direction, k=3)
table(knn.pred, Direction.2005)
mean(knn.pred==Direction.2005)
```
The results have improved slightly. But increasing K further turns out to provide no further improvements. It appears that for this data, QDA provides the best results of the methods that we have examined so far.

####4.6.6 An Application to Caravan Insurance Data

Finally, we will apply the KNN approach to the Caravan data set, which is part of the ISLR library. This data set includes 85 predictors that measure demographic characteristics for 5,822 individuals. The response variable is Purchase, which indicates whether or not a given individual purchases a caravan insurance policy. In this data set, only 6% of people purchased caravan insurance.

```{r}
dim(Caravan)
attach(Caravan)
summary(Purchase)
348/5822
```

Because the KNN classifier predicts the class of a given test observation by identifying the observations that are nearest to it, the scale of the variables matters. Any variables that are on a large scale will have a much larger effect on the distance between the observations, and hence on the KNN classifier, than variables that are on a small scale. For instance, imagine a data set that contains two variables, salary and age (measured in dollars and years, respectively). As far as KNN is concerned, a difference of $1,000 in salary is enormous compared to a difference of 50 years in age. Conse- quently, salary will drive the KNN classification results, and age will have almost no effect. This is contrary to our intuition that a salary difference of $1, 000 is quite small compared to an age difference of 50 years. Further- more, the importance of scale to the KNN classifier leads to another issue: if we measured salary in Japanese yen, or if we measured age in minutes, then we’d get quite different classification results from what we get if these two variables are measured in dollars and years. A good way to handle this problem is to standardize the datasothatall standardize
variables are given a mean of zero and a standard deviation of one. Then
all variables will be on a comparable scale. The scale() function does just scale() this. In standardizing the data, we exclude column 86, because that is the qualitative Purchase variable.

```{r}
standardized.X = scale(Caravan[,-86])
var(Caravan[,1])
var(standardized.X[,1])
var(standardized.X[,2])
```

We now split the observations into a test set, containing the first 1,000
observations, and a training set, containing the remaining observations. We fit a KNN model on the training data using K = 1, and evaluate its performance on the test data.

```{r}
test = 1:1000
train.X = standardized.X[-test,]
test.X = standardized.X[test,]
train.Y=Purchase[-test]
test.Y=Purchase[test]
set.seed(1)
knn.pred = knn(train.X, test.X, train.Y, k = 1)
mean(test.Y!=knn.pred)
mean(test.Y!="No")
```

The KNN error rate on the 1,000 test observations is just under 12%. At first glance, this may ap- pear to be fairly good. However, since only 6% of customers purchased insurance, we could get the error rate down to 6% by always predicting No regardless of the values of the predictors!

If the company tries to sell insurance to a random selection of customers, then the success rate will be only 6%, which may be far too low given the costs involved. Instead, the company would like to try to sell insurance only to customers who are likely to buy it. So the overall error rate is not of interest. Instead, the fraction of individuals that are correctly predicted to buy insurance is of interest.

Suppose that there is some non-trivial cost to trying to sell insurance
to a given individual. For instance, perhaps a salesperson must visit each potential customer. If the company tries to sell insurance to a random selection of customers, then the success rate will be only 6%, which may be far too low given the costs involved. Instead, the company would like to try to sell insurance only to customers who are likely to buy it. So the overall error rate is not of interest. Instead, the fraction of individuals that are correctly predicted to buy insurance is of interest.

```{r}
table(knn.pred, test.Y)
9/(68+9)
```
It turns out that KNN with K = 1 does far better than random guessing
among the customers that are predicted to buy insurance. Among 77 such customers, 9, or 11.7%, actually do purchase insurance. This is double the rate that one would obtain from random guessing.

Using K = 3, the success rate increases to 19%, and with K =5 the rateis 26.7%. This is over four times the rate that results from random guessing. It appears that KNN is finding some real patterns in a difficult data set!

```{r}
knn.pred = knn(train.X, test.X, train.Y, k = 3)
table(knn.pred, test.Y)
5/26

knn.pred = knn(train.X, test.X, train.Y, k = 5)
table(knn.pred, test.Y)
4/15
```

As a comparison, we can also fit a logistic regression model to the data.
If we use 0.5 as the predicted probability cut-off for the classifier, then we have a problem: only seven of the test observations are predicted to purchase insurance. Even worse, we are wrong about all of these! However, we are not required to use a cut-off of 0.5. If we instead predict a purchase any time the predicted probability of purchase exceeds 0.25, we get much better results: we predict that 33 people will purchase insurance, and we are correct for about 33% of these people. This is over five times better than random guessing!

```{r}
glm.fit = glm(Purchase~., data = Caravan, family=binomial,
              subset=-test)
glm.probs <- predict(glm.fit, Caravan[test,], type="response")
glm.pred = rep("No", 1000)
glm.pred[glm.probs>.5]="Yes"
table(glm.pred, test.Y)
0/7

glm.pred = rep("No", 1000)
glm.pred[glm.probs>.25]="Yes"
table(glm.pred, test.Y)
11/33
```

